What We Eval in the Shadows

### 2021-04-09 paper/main.tex
usage

\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
  \centering
  \includegraphics[width=75mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. There are \PkgEvalCallSites call sites to \eval in \PkgPackages
packages.  These represent \PkgPackagesRatio of all packages or \PkgCodeRatio
of the entire code size of CRAN. Over half of the packages have less than 3
\eval call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount
\eval calls), all packages contain less than \MaxEvalCallSitesRest \eval call
sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
\PkgFunsWithEval R functions (\CranFunsWithEvalRatio of all R functions in
CRAN).

For the dynamic analysis, we execute all the \CranRunnableScripts programs
extracted from \CranPackages packages. Any run that did not exercise one of the
\eval call sites in the packages was discarded. This left \packageNbruns runs of some
of the \packageCorpus packages to generate the data presented here. There were
\packageAllcalls invocations of \eval and its variants occurring in \packageTriggeredpkgs
packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
were hit at least once (\cf Fig~\ref{fig:traced-eval-callsites}). On average,
the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call sites,
which similar to the \PkgCodeCoverage of average package code
coverage\footnote{Not all packages with static call sites of \eval were
observed triggering that call site, this can be chalked down to incomplete code
coverage and occasional tracer failure (\PkgFailedProgramsRatio of programs do
not terminate successfully or within 30 minute timeout).}.  Figure~\ref{freq}
summarizes the frequency of dynamic calls to \eval with the left column being
the number of calls and the right, the number of packages in that range. There
are \packageFewcalls packages with low \eval frequency, fewer than 100 calls; and a
similar number of packages, \packageManycalls to be precise, that use \eval more than
1,000 times.  One package, \packageMaxcallspack, calls it \packageMaxcalls times and thus
accounts for over half of the observed calls.

\begin{figure}[b]

\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
\begin{minipage} {5cm}
  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
\tt 1 &--& \tt 10      & \packageBina  & \tt 1K &--&\tt 100K  & \packageBine\\
\tt 11 &--& \tt 100    & \packageBinb  & \tt 100K &--&\tt 1M  & \packageBinf\\
\tt 101 &--& \tt 1K    & \packageBinc  & \tt 1M &--&\tt 10M   & \packageBing\\
\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\hline
\end{tabular}
\caption{Call frequency}\label{freq}
\end{minipage}
&
\begin{minipage}{7cm}
\begin{tabular}{|@{\,}r|rrrr|}\hline
  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
           & & & \c{.parent} &\\\hline
\small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
\small Exercised sites&\packageTriggeredeval&\packageTriggeredevalq&\packageTriggeredevalparent&\packageTriggeredlocal\\
\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\hline
\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
\end{minipage}\end{tabular}\end{figure}

Figure~\ref{tab:variantseval} enumerate the use of variants of \eval. For each
of the four variants, the first rows shows the number of call sites in the
corpus (\emph{static}), the second row is the number of call sites that were
encountered during analysis (\emph{exercised}), the last row is the number of
calls (\emph{invocations}). The overwhelming majority of call sites and dynamic
calls, are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
\c{local} are barely used at all. The difference between sites and exercised
sites underscores the limitations due to code coverage.

\begin{wrapfigure}{4}{5.8cm}
\vspace{-4mm}
\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
\tt 51 &--& \tt 100  & \packageRunbinb & \tt 1001 &--& \tt 1500  & \packageRunbinf\\
\tt 101 &--& \tt 250 & \packageRunbinc & \tt 1501 &--& \tt 2000  & \packageRunbing\\
\tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\hline
\end{tabular}

@@

  \includegraphics[width=5.4cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}

@@

\vspace{-2mm} \caption{Normalized invocations} \label{cn}\vspace{-2mm}
\end{wrapfigure}

Figure~\ref{cn}(a) shows normalized invocation counts per call site; on the left
are average number of calls from a given site and given run, on the right are
counts of sites that fall in that range. For instance, \packageRunbinh sites are
invoked on average over 2,000+ times per run. Larger numbers suggest loops or
recursive contexts -- this seems to be the exception as most \evals, \packageRunbina,
are exercised 50 times or less. Figure~\ref{cn}(b) zooms in on low frequency
call sites. The x-axis shows normalized invocations and the y-axis is the number
sites for that value. Most low-frequency \evals are invoked only once, about
half as many are invoked twice; after that the frequency quickly drops.

The \eval can be passed any value, but if the argument is not an expression,
\eval returns it without doing anything. Expressions account for \packageCodepercent of
arguments in our corpus. More specifically, \packageSymbolpercent of arguments are
\c{symbol}s (single variables such as \c{x}), \packageLanguagepercent are \c{language}
object (a single expression, such as \c{x+1}), and \packageExpressionpercent are
\c{expression} objects (multiple statements). Further inspection reveals that
most symbols, \packageGgplotsymbolpercent to be exact, come from a single call in the
\c{ggplot2} package and have have the value \c{\_inherit}. Even if we remove
this outlier, the majority of \evals are simple expressions.

To estimate how much executable code is injected through \eval, we measure the
the number of abstract syntax tree nodes as returned by the parser; for example,
\c{x+1} counts as 3.\footnote{We attempted to measure string lengths after
de-parsing expressions, but these measurements were dominated by de-parsed data
objects, which could range in the MBs.} The median argument size is
\packageMedianszeval (due to the many symbols) and the average is \packageAvgszeval nodes. The
largest \eval input observed is \packageMaxszeval\,-- a significant chunk of code.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
of fewer or equal to 25 nodes.
The x-axis is the size of arguments in number of
nodes, and the y-axis is the count of arguments with that size. The size drops
rapidly, with few observations larger than 15 nodes. The long tail is omitted
from the graph for legibility.
%
\begin{wrapfigure}{r}{5.8cm}
 \hspace*{-10mm}
 \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
 \caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}
%

To estimate the work performed in \evals, we count the instructions executed by
interpreter. Most invocations perform relatively little work with
\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin plot of
Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50 instructions, it is
dominated by trivial symbol look ups. Figure~\ref{ev}(b) has the work-intensive
\evals which go all the way to \packageMaxeventsRnd instructions.

\begin{figure}[tb!]
\begin{tabular}{@{}c@{}c@{}}
\begin{minipage}{7.5cm}
 \includegraphics[width=\textwidth]{events_per_pack_small}
\end{minipage}&\begin{minipage}{7.5cm}
  \includegraphics[width=\textwidth]{events_per_pack_large}
\end{minipage}\\[-3mm]
\small (a) Small & \small (b) Large
\end{tabular}
 \caption{Instructions per \eval} \label{ev}
\end{figure}

\paragraph{Base}
We recorded \baseAllcalls \eval calls in the base libraries with fewer than 4\%
of those being to any variants. Unlike CRAN, Base calls \c{evalq} (\baseEvalqs).
Out of \baseStaticeval call sites, \baseTriggeredevalpct are exercised. Most
arguments, \baseCodepercent, are expressions. However, unlike CRAN,
\baseLanguagepercent of the arguments are language objects. The median argument
size is \baseMedianszeval, which is more than for CRAN. The maximum size is
\baseMaxszeval. Small instruction counts, less than 50, amount for
\baseSmalleventspct of calls.
\todo{More info about base and conclude it here.}

\begin{figure}[tb] \centering
  \includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf} \centering
  \caption{\eval call sites coverage of the \PkgPackages packages.}%
  \label{fig:traced-eval-callsites}
\end{figure}

\paragraph{Kaggle}
Only \kaggleNbruns scripts trigger \eval. In total, \kaggleAllcalls \eval were
recorded; none are calls to variants. Out of \kaggleStaticeval call sites,
\kaggleTriggeredeval were exercised. Most call sites are invoked only once. Only
one site is called \kaggleMaxcalls times. Expressions amount for
\kaggleCodepercent of arguments. Unlike CRAN corpus, there are very few symbols
(\kaggleSymbolpercent). Most expressions result from calls to \c{parse}, thus
most \evals start with strings. The median argument size is \kaggleMedianszeval,
Which makes sense, as few arguments are only symbols. The largest argument is
\kaggleMaxszeval. The distribution of instructions per eval is similar to CRAN;
\kaggleSmalleventspct of \evals execute fewer than 50 instructions.
\todo{More info about kaggle and conclude it here.}

\paragraph{Discussion}
These results show that \eval is central to the language implementation as it is
omnipresent in the relatively small Base library. Packages, which are developed
by experienced programmers, make regular and varied use of the facility.
Finally, scripts are written by less sophisticated users, and likely perform
simpler tasks, and thus have lesser need for \eval, and its uses originate from
strings which are easier to manipulate for end-users.

@@

\paragraph{Base}
\todo{I would remove this - we conclude Base in 4}
The spread of forms in Base are similar to CRAN with a larger share of \c{V}, as
it is present in \baseMinimizedpropsitesa\% of the sites and \c{X} is a distant
second with \baseMinimizedpropsitesb\%.

\paragraph{Kaggle}
\todo{I would remove this - we conclude Kaggle in 4}
Kaggle is dominated by \c{V}, with \kaggleMinimizedpropsitesa\% of the
\kaggleNbCallSites sites.

>>>>

\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
  \centering
  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. We use the word \emph{site} to an occurrence of a call site to the \eval
function in the source code, and \emph{call} to denote an observed invocation of
the \eval function.

\subsection{CRAN}

There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
of packages calling \eval is thus \PkgPackagesRatio. Over half of these packages
have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage which
has \MaxEvalCallSitesCount sites, all packages contain fewer than
\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist} which
shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
functions (\CranFunsWithEvalRatio of all functions in CRAN).

\begin{figure}[!b]
\small
\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
\begin{minipage} {5cm}
  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
\tt 1 &--& \tt 10      & \packageBina  & \tt 1K &--&\tt 100K  & \packageBine\\
\tt 11 &--& \tt 100    & \packageBinb  & \tt 100K &--&\tt 1M  & \packageBinf\\
\tt 101 &--& \tt 1K    & \packageBinc  & \tt 1M &--&\tt 10M   & \packageBing\\
\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\hline
\end{tabular}
\caption{Call frequency}\label{freq}
\end{minipage}
&
\begin{minipage}{7cm}
\begin{tabular}{|@{\,}r|rrrr|}\hline
  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
           & & & \c{.parent} &\\\hline
\small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
\small Exercised sites&\packageTriggeredeval&\packageTriggeredevalq&\packageTriggeredevalparent&\packageTriggeredlocal\\
\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\hline
\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
\end{minipage}\end{tabular}

%\end{figure}

%\begin{figure}[b] \centering
  \includegraphics[width=.78\textwidth]{traced-eval-callsites.pdf} \centering
  \caption{\eval call sites coverage of the \PkgPackages packages.}%
  \label{fig:traced-eval-callsites}
\end{figure}

For dynamic analysis, we run \CranRunnableScripts programs extracted from
\CranPackages packages. Any run that does not exercise \eval is discarded. This
left \packageNbruns runs from \packageCorpus packages. There were
\packageAllcalls calls in \packageTriggeredpkgs packages originating from
\PkgHitEvalCallSites unique sites. In terms of coverage, the data exercised
\PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
metric for the packages which is \PkgCodeCoverage. The fact that not all sites
are exercised can be can be chalked down to incomplete tests occasional analysis
failures (\PkgFailedProgramsRatio of programs crashed or timed out).
Fig~\ref{fig:traced-eval-callsites} shows the number of sites that were
exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
dynamic calls to \eval with the left column being the number of calls and the
right, the number of packages in that range. There are \packageFewcalls packages
with low \eval frequency, fewer than 100 calls; and a smaller, but still
significant, number of packages, \packageManycalls to be precise, that use \eval
more than 1,000 times. Package \packageMaxcallspack makes \packageMaxcalls calls
and thus accounts for over half of the observations.
Figure~\ref{tab:variantseval} summarizes the use of variants of \eval. For each
of the four variants, the first rows shows the number of sites in the corpus
(\emph{static}), the second row is the number of sites that were encountered
during analysis (\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The overwhelming majority of sites and calls, are to \eval
itself, \c{eval.parent} is rare, and both \c{evalq} and \c{local} are barely
used at all. The difference between sites and exercised sites underscores the
limitations due to code coverage.

\begin{wrapfigure}{4}{5.8cm}
\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
\tt 51 &--& \tt 100  & \packageRunbinb & \tt 1001 &--& \tt 1500  & \packageRunbinf\\
\tt 101 &--& \tt 250 & \packageRunbinc & \tt 1501 &--& \tt 2000  & \packageRunbing\\
\tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\hline
\end{tabular}

@@

  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}

@@

\caption{Normalized calls} \label{cn}\vspace{-2mm}

\medskip
\medskip

\begin{tabular}{c}
  {\hspace{-3cm}\includegraphics[width=0.8\textwidth]{size_loaded_distribution}}
\end{tabular}
 \caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}

Figure~\ref{cn}(a) shows normalized call counts per site; on the left are
average number of calls from a given site and given run, on the right are counts
of sites that fall in that range. For instance, \packageRunbinh sites are
invoked 2,000+ times per run. Larger numbers suggest loops or recursive contexts
-- this seems to be the exception as most \evals, \packageRunbina, are exercised
50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
shows normalized calls and the y-axis is the number sites for that value. Most
low-frequency \evals are invoked only once, about half as many are invoked
twice; after that the frequency quickly drops.

\Eval takes any value, but if its argument is not an expression, \eval returns
it unchanged. Expressions account for \packageCodepercent of arguments in our
corpus. More specifically, \packageSymbolpercent are \c{symbol}s (single
variables such as \c{x}), \packageLanguagepercent are \c{language} object (a
single expression, such as \c{x+1}), and \packageExpressionpercent are
\c{expression} objects (multiple statements). Further inspection reveals that
most symbols, \packageGgplotsymbolpercent to be exact, come from a single call
in the \c{ggplot2} package and have have the value \c{\_inherit}. Even if we
remove this outlier, the majority of \evals are simple expressions.

To estimate how much executable code is injected through \eval, we measure the
the number of nodes in the expressions; for example, \c{x+1} counts as 3. We
also measured string lengths of unparsed expressions, but these measurements
were dominated by the size of data objects which could range in the MBs. The
median argument size is \packageMedianszeval (due to the symbols) and the
average is \packageAvgszeval nodes. The largest \eval input observed is
\packageMaxszeval\,-- a significant chunk of code.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
nodes, and the y-axis is the count of arguments with that size. The size drops
rapidly, with few observations larger than 15 nodes. The long tail is omitted
for legibility.

To estimate the work performed in \evals, we count instructions executed by
interpreter. Most invocations perform relatively little work with
\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin
plot of Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50
instructions, it is dominated by trivial symbol look ups. Figure~\ref{ev}(b) has
the work-intensive \evals which go all the way to \packageMaxeventsRnd
instructions.

\begin{figure}[tb!]
\begin{tabular}{@{}c@{}c@{}}
\begin{minipage}{7.5cm}
 \includegraphics[width=\textwidth]{events_per_pack_small}
\end{minipage}&\begin{minipage}{7.5cm}
  \includegraphics[width=\textwidth]{events_per_pack_large}
\end{minipage}\\[-3mm]
\small (a) Small & \small (b) Large
\end{tabular}
 \caption{Instructions per call} \label{ev}
\end{figure}

\subsection{Base}
We recorded \baseAllcalls \eval calls in the base libraries with fewer than 4\%
of those being to any variants (this was done on a small subset of the whole
corpus). Unlike CRAN, Base calls \c{evalq} (\baseEvalqs). Out of \baseStaticeval
call sites, \baseTriggeredevalpct are exercised. Most arguments,
\baseCodepercent, are expressions. However, unlike CRAN, \baseLanguagepercent of
the arguments are language objects. The median argument size is
\baseMedianszeval, which is more than for CRAN. The maximum size is
\baseMaxszeval. Small instruction counts, less than 50, amount for
\baseSmalleventspct of calls. The challenge dealing with Base is that every
program uses it and it can thus generate extreme amounts of data but the data is
quite predictable. For the rest of the paper, we only mention base when there
are surprising observations.

\medskip

\subsection{Kaggle}
Only \kaggleNbruns scripts trigger \eval. In total, \kaggleAllcalls \eval were
recorded; none are calls to variants. Out of \kaggleStaticeval call sites,
\kaggleTriggeredeval were exercised. Most call sites are invoked only once. Only
one site is called \kaggleMaxcalls times. Expressions amount for
\kaggleCodepercent of arguments. Unlike CRAN corpus, there are very few symbols
(\kaggleSymbolpercent). Most expressions result from calls to \c{parse}, thus
most \evals start with strings. The median argument size is \kaggleMedianszeval,
Which makes sense, as few arguments are only symbols. The largest argument is
\kaggleMaxszeval. The distribution of instructions per eval is similar to CRAN;
\kaggleSmalleventspct of \evals execute fewer than 50 instructions. Manual
inspection of \eval usage in Kaggle suggest that it is consistent with the data
obtained for CRAN. We do not discuss it further.

\medskip

\subsection{Discussion}
These results show that \eval is central to the language implementation as it is
omnipresent in the relatively small Base library. The Base library is, in turn,
used by every single package. To avoid having Base dominate the result we have
excluded calls occurring there from the data we report when analyzing packages.
CRAN Packages, which are typically developed by experienced programmers, make
regular and varied use of \eval. They represent our most interesting data set as
these packages are the result of over 20 years of contributions by thousands of
authors. The code in packages is well maintained and relatively well tested.
Finally, Kaggle scripts are often written by less sophisticated users, and
likely perform simpler tasks, and thus have lesser need for \eval, and its uses
originate from strings as these are easier to manipulate for end-users. Due to
the relative paucity of data in Kaggle, we do not pursue the data set further.

@@
